Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-07 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing typo, mainly for commit message testing
small change to test final commits
Section 3 Keynote: turn URLs into a references

\todo{change references to papers or presentations to cites}

@@

The first
keynote~\cite{WSSSPE1-keynote1}
was delivered by Philip E. Bourne of University of California, San Diego.  Bourne is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and
PDB~\cite{PDB}.  He is
working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and would serve as a key function of software in biomedical sciences.

Bourne's keynote presentation is entiled ``A Recipe for Sustainable Software''.
He emphasized that sustainability for software ``does not just
mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the App Store model. He mentioned
BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
as distribution channels.  On a related note, Bourne observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

@@

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
good will only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology
transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}

@@

He then presented the usual arguments on directly valuing software as a
research output alongside papers. Readers will be familiar with this
debate. An interesting reference here is ``What
does peer review mean when applied to computer code?, Rosemary Dickin.''
\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}}

@@

He asked, of this practice, ``why didn't we know better?'' and argued
this was because we were taught to focus on immediate research
outcomes and not on continuously improving and building on tools for
research. He then asked, later in our working lives, when we do know
better, why don't we act any different, and argued this was due to
incentives and their lack: only the immediate products of research,
not the software, are valued.  He referenced Victoria Stodden's talk
at
OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
which he said argued these points well.

He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov
Chain Monte Carlo code by Dan Forman
Mackey\footnote{\url{http://dan.iel.fm/}} and Joe
Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}
as good examples of scientific projects using GitHub.

C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with
regard to reusable software, ``we should just start doing it.'' Arfon
noted that documentation should be ``treated as a first class
entity.''  He noted that some software fields have excellent cultures
of module reuse, such as
rubygems\footnote{\url{http://rubygems.org/}},
PyPI\footnote{\url{https://pypi.python.org/pypi}}, and
CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
ultra-low-friction collaboration through these repositories. This has
not happened in highly numerical, compiled language scientific
software.

>>>>

@@

The first
keynote~\cite{WSSSPE1-keynote1}
was delivered by Philip E. Bourne of University of California, San Diego.  Bourne is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and
PDB~\cite{pdb-web}.  He is
working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and would serve as a key function of software in biomedical sciences.

Bourne's keynote presentation is entiled ``A Recipe for Sustainable Software''.
He emphasized that sustainability for software ``does not just
mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the App Store model. He mentioned
BioJava~\cite{biojava-web} and
OpenScienceDataCloud~\cite{osdc-web}
as distribution channels.  On a related note, Bourne observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

@@

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
good will only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology transfer.~\cite{sbir-web, fased-web}.

@@

He then presented the usual arguments on directly valuing software as a
research output alongside papers. Readers will be familiar with this
debate. An interesting reference here is~\cite{peer-review-code}.

@@

He asked, of this practice, ``why didn't we know better?'' and argued
this was because we were taught to focus on immediate research
outcomes and not on continuously improving and building on tools for
research. He then asked, later in our working lives, when we do know
better, why don't we act any different, and argued this was due to
incentives and their lack: only the immediate products of research,
not the software, are valued.  He referenced Victoria Stodden's talk
at OKCon~\cite{okcon-stodden-talk}
which he said argued these points well.

He cited the EMCEE Markov
Chain Monte Carlo code~\cite{emcee-web} by Dan Foreman-Mackey and contributors
as good examples of scientific projects using GitHub.

C. Titus Brown~\cite{ged-web} argues that with
regard to reusable software, ``we should just start doing it.'' Arfon
noted that documentation should be ``treated as a first class
entity.''  He noted that some software fields have excellent cultures
of module reuse, such as
RubyGems~\cite{rubygems-web},
PyPI~\cite{pypi-web}, and
CPAN~\cite{cpan-web}, where there is effectively
ultra-low-friction collaboration through these repositories. This has
not happened in highly numerical, compiled language scientific
software.
